Topology and geometry of the dark matter web: a multistream view

### 2018-06-12 Draft/Chapter4/Paper2017a_Nobmath.tex
spell checked

Heuristic analysis of the multistream field has been done in Chapter \ref{chapter3} where the dark matter structures were classified based on local multistream variational approach. In this chapter, we analyse the dark matter structures based on their phyiscal shapes and connectivity. Extensive studies of geometrical and topological properties in cosmic matter density field have been done in the past. Here we extend some of these studies into the Lagrangian sub-manifold regime -- taking advantage of certain unique properties of the multistream field supplement our understanding of dark matter structures. The analysis and text in this chapter overlaps with \cite{Ramachandra2017}.

@@

A related but different question is how to select individual non-linear structure, like halos, filaments and walls by using a local parameter. In particular the density threshold has been  used on numerous occasions especially for identifying halos and voids. As a rule the choices of particular values have not been justified by solid physical evidences. The virial mass and virial radius of a halo are often used as direct indicators of gravitationally bound objects but they are  determined  by a nonlocal quantity -- the mean overdensity of the halo. 
An interesting compaison of several kinds of boundaries of halos was provided by \cite{More2015}. In particular they considered  the virial radius $R_{\rm vir}$,  $R_{\rm  200m}$,  the splashback radius $R_{\rm sp}$, and $R_{\rm infall}$. The splashback
radius is defined as an average distance from the center of the halo to the most external caustic if it was resolved. The authors argue
that it is ``a more physical halo boundary choice" than ``commonly defined to enclose a density contrast $\Delta_{\rm m,c}$ relative to 
a reference (mean or critical) density.
This is the boundary where the number of streams falls from three to one in the multistream field.

@@

The rest of the paper is organised as follows: we describe the cosmological simulations in Section \ref{sec:simulation}. Some of the important features of the multistream field are described in Section \ref{sec:multiCalc}. Topology of the single-streaming voids is discussed in \ref{sec:voids} and that of the multistream structure is investigated using percolation theory in Section \ref{sec:percolation}. . Discussion of the local geometry of multistream field using Hessian matrices is done in Section \ref{sec:hessian}.

@@

We studied the $n_{str}$ environment of the haloes detected using the Friends-of-Friends method (FOF-\citealt{Davis1985}) as illustrated in \autoref{fig:Vfr_all}. FOF-haloes with more than 20 particles are detected using linking-length of $b=0.2$ in the simulation with $128^3$ particles. We implement the diagnosis method prescribed in \cite{Ramachandra2015}: a large number of points are randomly selected on diagnostic spherical surfaces centred at the FOF-centre of the halo. Multi-stream values are iteratively calculated at these spherical surfaces of various radii. We define the distance from centre of a halo, $r_{10}$, where $n_{str} = 1$ at 10 per cent of the surface of the diagnostic sphere. Distribution of this void-distance parameter is compared to the virial radii $r_{vir}$ of the FOF-haloes. Surprisingly, $r_{10}$ distribution peaks at slightly lower values than the $r_{vir}$ distribution. This implies a large number of FOF-haloes are in the vicinity of the void.

@@

The eigenvalues of ${\bf H}(-n_{str})$ span a large range of values in our cosmological simulation. The largest eigenvalue of the triplets, $\lambda_1$ having large positive values throughout the multistream web structure ( see \autoref{fig:evals123} ). Absolute values $|\lambda_1|$, $|\lambda_2|$ and $|\lambda_3|$ peak around the neighbourhood of intersections of filaments. These junctions are usually high streaming regions due to shell crossing from multiple directions. \cite{Ramachandra2015} observed that these regions with intersecting filaments are in the vicinity of large FOF haloes.

@@

If the Hessian matrices are positive definite in a region, i.e., if all the eigenvalues are strictly positive, then the interior of this convex region has at-most one minimum. For our choice of $-n_{str}(\bf{x})$ as the domain of Hessian, this means that the convex neighbourhoods around local maxima of the multistream field are isolated by the positive definite Hessian matrices. Closed surface contours at high streaming or the most non-linear regions are selected. These regions my indeed be the regions of dark matter haloes.

@@

Hessian eigenvalues are generally defined on continuous functions. Although our domain of the Hessian is an inhererntly integer-valued field, it describes the multistream structure at the level of diagnostic grid. Hence it may be considered to be numerically equivalent to a continuous function where the numerical approxiamtion of differentation is a valid operation. This can be verified mathematically by finding that Hessian ${\bf H}(-n_{str})$ is symmetric (Appendix \ref{appendix:Eigen} shows the numerical approximation of the Hessian matrix term for generic unfiltered multistream field.)

Smoothing the multistream field (at the refinement level of $l_l/l_d=$ $1$ or $2$) effectively reduces noise. There is also a systematic variation in the distribution of smoothed $n_{str}$ values as shown in \autoref{fig:nstrSmooth}. Volume fraction of the single-streaming voids only varies from $90.8$ per cent without smoothing to $89.1$ per cent for the Gaussian softening length of 0.39 $h^{-1} Mpc$ (twice the length of diagnostic grid $l_d$). On the other hand, $n_{str} = 3$ regions gain volume fraction from $4.9$ per cent in un-smoothed field to $7.1$ per cent for 0.39 $h^{-1} Mpc$. This is seen in the multistream structures of smootheing scales of 0.39 $h^{-1} Mpc$ in \autoref{fig:nstrSmoothSmall}. Multi-stream regions with $ 3 < n_{str} \leq 100 $ occupy correspondingly lower volumes for higher smoothing, and the variation is noisy beyond $n_{str} > 100$. \autoref{fig:nstrSmoothSmall} shows the multistream field on a small slice of the simulation at different softening scales, and walls and filaments are resolved better with increasing softening.

@@

Formation of multiple velocity streams in the context of structure formation has been known in the past, starting from Zel'dovich approximation. Quantification of the multistreams in N-body simulations, however, was recently achieved by \cite{Shandarin2012} and \cite{Abel2012} using the Lagrangian sub-manifold. In our study, the multistream fields are calculated using the tessellation algorithm by \cite{Shandarin2012}. We have analysed, for the first time, the local geometry and percolation properties of the cosmic web using this multistream field.

Distinguishing the configuration space into void and non-void is one of the uses of the multistream field. Lagrangian sub-manifold has no folds in the beginning, thus $n_{str} = 1$ uniformly throughout the simulation. Gravitational instability folds the sub-manifold in complicated ways, however, most of the volume has particles without any collapse. \cite{Shandarin2012} and \cite{Ramachandra2015} observed that the single-streaming voids occupy around 85-90 per cent of the simulations at $z=0$. In this study, we found that the void regions are also connected in a way that the largest percolating void occupies more than 99 per cent of the all the single-streaming regions. Recent study by \cite{Wojtak2016a} uses a watershed transform method in the density field prescribed by Lagrangian tessellations (\citealt{Shandarin2012} and \citealt{Abel2012}) to analyse the evolution of isolated voids. Another recent study by \cite{Falck2015} on ORIGAMI-voids also reveal a similar percolation at the limit of simulation resolution. They observed persistence of this phenomenon for different resolutions of the N-body simulation. Multi-stream analysis, on the other hand, is not limited to mass resolution of the simulation. Our multistream analysis refined upto 8 times the simulations resolution revealed that the percolation phenomenon still persists. However, at high refinements of the multistream field, we observed small voids that are enclosed by highly resolved non-void membranes.

@@

Walls are the first collapsed structures in the dark matter Universe. At highly refined multistream field, thin membranes of the structures are often resolved, revealing small voids enclosed by them (compare two top panels in \autoref{fig:NstrCaust}). These preliminary structures are separated from the voids by caustic surfaces. These caustics have volume measure zero, which makes detection of their surface harder in the multistream field, even at very high resolutions. On the other hand, caustic surfaces themselves can by detected using the Lagrangian sub-manifold by identifying the common faces of neighbouring tetrahedra 
with opposite volume signs \citep{Shandarin2012}. They are shown in the bottom panel in  \autoref{fig:NstrCaust}. One can see that increasing the refinement factor from 2 to 8 adds mostly walls but the complete wall structure shown in the bottom panel is still considerably greater. Please note that the plots in  two top panels adjusted  exactly to the simulation box in Eulerian space, and the bottom plot shows the Lagrangian box mapped to Eulerian space without adjusting to the simulation box.

There are extensive number of topological indicators in the context of density fields or spatial co-ordinates - such as alpha shapes, Betti numbers, genus statistics. Although a comparitive study of these topolgical measures in multistream fields may be interesing, it is not the intent of this chapter (See Chapter \ref{chapter3b} for a detailed comparison of various structure finding algorithms). In this study, we only investigate percolation transitions in excursion sets of multistreams as a preliminary analysis of topological connectivities. Excursion sets in density fields are shown to have quick percolation transitions \citep{Shandarin2010b} and a similar trend in multistream field is investigated here.

@@

Since the $n_{str}$ field in this study is calculated on regular grids, the boudaries of the structures are not exactly traced. Outlining foldings in the Lagrangian sub-manifolds exactly as shown in \autoref{fig:NstrCaust} or in the {\it flip-flop} calculations shown in \cite{Shandarin2016} give point datasets which are considerably more difficult to analyze. However, recent advancements in computational topology - such as the adaptation of the watershed transforms (using SpineWeb -\citealt{Aragon-Calvo2008} and Morse theory (using DisPerSe - \citealt{Sousbie2011e} and Felix - \citealt{Shivshankar2015a}) to inherently discrete datasets may be useful in the topological analyses of flip-flop fields and caustics.

The multistream field is a scalar function of Eulerian coordinates. We have analysed functional variation of the $-n_{str}(\bf{x})$ field using Hessian eigenvalues. The Hessian analysis is generally done for inherently continuous fields, For example, Hessian analysis has been previously studied for smoothed density fields (see \citealt{Sousbie2008c}, \citealt{Aragon-Calvo2007}, \citealt{Aragon-Calvo2010}, \citealt{Cautun2014a} etc.), gravitational potential and velocity shear tensor (\citealt{Hoffman2012a}, \citealt{Libeskind2013}, \citealt{Hahn2007}, \citealt{Forero-Romero2009a}, \citealt{Hoffman2012a} and \citealt{Cautun2014a}). Although the multistream field has discrete values by definition, it may be considered smooth for numerical analysis at the scale of grid length of the field. The resulting Hessian eigenvalues characterize the geometry in a four-dimensional hyper-space of ($-n_{str}, x, y, z$). The boundary of a region with $\lambda_1 \geq \lambda_2 \geq \lambda_3 > 0$ is a closed convex contour in this hyper-space, and thus it's projection onto the three-dimensional Lagrangian space is also closed and convex.

@@

Dark matter haloes, being localised structures, are uniquely convenient for our local Hessian analysis. Conditions of $\lambda_1 > 0 > \lambda_2 \geq \lambda_3 $ and $\lambda_1 \geq \lambda_2 > 0 > \lambda_3 $ also give information about curvature. Hessian eigenvalue analysis at high resolution of multistream fields may be very interesting in understanding the tubular edges of filaments and surfaces of walls at smaller scales.

@@

\item We use the multistream field as a proxy for distinguishing of the DM web from DM voids: the web is defined as the regions
with number of streams greater than one and thus voids  as  a single stream regions. The boundary between them  
representing a sharp transition from one-- to three-- stream flow regions
would be a caustic surface in the density field if the mass and spatial resolutions were sufficiently high. They were clearly seen in 
2D simulations by \cite{Melott1989} as well as in 3D simulations by \cite{Angulo2016}, \cite{Hahn2016a}, \cite{Hahn2013} and in velocity fields \cite{Hahn2015a}.

@@

The Lagrangian sub-manifold contains dynamical information of structure formation. We analysed the multistream field that contains the information of foldings in the sub-manifold.  Connectivities in the void and non-void components of the multistream web reveal several details about structure of the Universe that are not probed by traditional density fields. In addition, we demonstrated the use of geometrical features of the multistream field in identifying potential dark matter halo candidates in cosmological N-body simulations.

>>>>

Heuristic analysis of the multistream field has been done in Chapter \ref{chapter3} where the dark matter structures were classified based on local multistream variational approach. In this chapter, we analyze the dark matter structures based on their physical shapes and connectivity. Extensive studies of geometrical and topological properties in cosmic matter density field have been done in the past. Here we extend some of these studies into the Lagrangian sub-manifold regime -- taking advantage of certain unique properties of the multistream field supplement our understanding of dark matter structures. The analysis and text in this chapter overlaps with \cite{Ramachandra2017}.

@@

A related but different question is how to select individual non-linear structure, like halos, filaments and walls by using a local parameter. In particular the density threshold has been  used on numerous occasions especially for identifying halos and voids. As a rule the choices of particular values have not been justified by solid physical evidences. The virial mass and virial radius of a halo are often used as direct indicators of gravitationally bound objects but they are  determined  by a nonlocal quantity -- the mean over-density of the halo. 
An interesting comparison of several kinds of boundaries of halos was provided by \cite{More2015}. In particular they considered  the virial radius $R_{\rm vir}$,  $R_{\rm  200m}$,  the splashback radius $R_{\rm sp}$, and $R_{\rm infall}$. The splashback
radius is defined as an average distance from the center of the halo to the most external caustic if it was resolved. The authors argue
that it is ``a more physical halo boundary choice" than ``commonly defined to enclose a density contrast $\Delta_{\rm m,c}$ relative to 
a reference (mean or critical) density.
This is the boundary where the number of streams falls from three to one in the multistream field.

@@

The rest of the paper is organized as follows: we describe the cosmological simulations in Section \ref{sec:simulation}. Some of the important features of the multistream field are described in Section \ref{sec:multiCalc}. Topology of the single-streaming voids is discussed in \ref{sec:voids} and that of the multistream structure is investigated using percolation theory in Section \ref{sec:percolation}. . Discussion of the local geometry of multistream field using Hessian matrices is done in Section \ref{sec:hessian}.

@@

We studied the $n_{str}$ environment of the haloes detected using the Friends-of-Friends method (FOF-\citealt{Davis1985}) as illustrated in \autoref{fig:Vfr_all}. FOF-haloes with more than 20 particles are detected using linking-length of $b=0.2$ in the simulation with $128^3$ particles. We implement the diagnosis method prescribed in \cite{Ramachandra2015}: a large number of points are randomly selected on diagnostic spherical surfaces centered at the FOF-center of the halo. Multi-stream values are iteratively calculated at these spherical surfaces of various radii. We define the distance from center of a halo, $r_{10}$, where $n_{str} = 1$ at 10 per cent of the surface of the diagnostic sphere. Distribution of this void-distance parameter is compared to the virial radii $r_{vir}$ of the FOF-haloes. Surprisingly, $r_{10}$ distribution peaks at slightly lower values than the $r_{vir}$ distribution. This implies a large number of FOF-haloes are in the vicinity of the void.

@@

The eigenvalues of ${\bf H}(-n_{str})$ span a large range of values in our cosmological simulation. The largest eigenvalue of the triplets, $\lambda_1$ having large positive values throughout the multistream web structure ( see \autoref{fig:evals123} ). Absolute values $|\lambda_1|$, $|\lambda_2|$ and $|\lambda_3|$ peak around the neighborhood of intersections of filaments. These junctions are usually high streaming regions due to shell crossing from multiple directions. \cite{Ramachandra2015} observed that these regions with intersecting filaments are in the vicinity of large FOF haloes.

@@

If the Hessian matrices are positive definite in a region, i.e., if all the eigenvalues are strictly positive, then the interior of this convex region has at-most one minimum. For our choice of $-n_{str}(\bf{x})$ as the domain of Hessian, this means that the convex neighborhoods around local maxima of the multistream field are isolated by the positive definite Hessian matrices. Closed surface contours at high streaming or the most non-linear regions are selected. These regions my indeed be the regions of dark matter haloes.

@@

Hessian eigenvalues are generally defined on continuous functions. Although our domain of the Hessian is an inherently integer-valued field, it describes the multistream structure at the level of diagnostic grid. Hence it may be considered to be numerically equivalent to a continuous function where the numerical approximation of differentiation is a valid operation. This can be verified mathematically by finding that Hessian ${\bf H}(-n_{str})$ is symmetric (Appendix \ref{appendix:Eigen} shows the numerical approximation of the Hessian matrix term for generic unfiltered multistream field.)

Smoothing the multistream field (at the refinement level of $l_l/l_d=$ $1$ or $2$) effectively reduces noise. There is also a systematic variation in the distribution of smoothed $n_{str}$ values as shown in \autoref{fig:nstrSmooth}. Volume fraction of the single-streaming voids only varies from $90.8$ per cent without smoothing to $89.1$ per cent for the Gaussian softening length of 0.39 $h^{-1} Mpc$ (twice the length of diagnostic grid $l_d$). On the other hand, $n_{str} = 3$ regions gain volume fraction from $4.9$ per cent in un-smoothed field to $7.1$ per cent for 0.39 $h^{-1} Mpc$. This is seen in the multistream structures of smoothing scales of 0.39 $h^{-1} Mpc$ in \autoref{fig:nstrSmoothSmall}. Multi-stream regions with $ 3 < n_{str} \leq 100 $ occupy correspondingly lower volumes for higher smoothing, and the variation is noisy beyond $n_{str} > 100$. \autoref{fig:nstrSmoothSmall} shows the multistream field on a small slice of the simulation at different softening scales, and walls and filaments are resolved better with increasing softening.

@@

Formation of multiple velocity streams in the context of structure formation has been known in the past, starting from Zel'dovich approximation. Quantification of the multistreams in N-body simulations, however, was recently achieved by \cite{Shandarin2012} and \cite{Abel2012} using the Lagrangian sub-manifold. In our study, the multistream fields are calculated using the tessellation algorithm by \cite{Shandarin2012}. We have analyzed, for the first time, the local geometry and percolation properties of the cosmic web using this multistream field.

Distinguishing the configuration space into void and non-void is one of the uses of the multistream field. Lagrangian sub-manifold has no folds in the beginning, thus $n_{str} = 1$ uniformly throughout the simulation. Gravitational instability folds the sub-manifold in complicated ways, however, most of the volume has particles without any collapse. \cite{Shandarin2012} and \cite{Ramachandra2015} observed that the single-streaming voids occupy around 85-90 per cent of the simulations at $z=0$. In this study, we found that the void regions are also connected in a way that the largest percolating void occupies more than 99 per cent of the all the single-streaming regions. Recent study by \cite{Wojtak2016a} uses a watershed transform method in the density field prescribed by Lagrangian tessellations (\citealt{Shandarin2012} and \citealt{Abel2012}) to analyze the evolution of isolated voids. Another recent study by \cite{Falck2015} on ORIGAMI-voids also reveal a similar percolation at the limit of simulation resolution. They observed persistence of this phenomenon for different resolutions of the N-body simulation. Multi-stream analysis, on the other hand, is not limited to mass resolution of the simulation. Our multistream analysis refined up to 8 times the simulations resolution revealed that the percolation phenomenon still persists. However, at high refinements of the multistream field, we observed small voids that are enclosed by highly resolved non-void membranes.

@@

Walls are the first collapsed structures in the dark matter Universe. At highly refined multistream field, thin membranes of the structures are often resolved, revealing small voids enclosed by them (compare two top panels in \autoref{fig:NstrCaust}). These preliminary structures are separated from the voids by caustic surfaces. These caustics have volume measure zero, which makes detection of their surface harder in the multistream field, even at very high resolutions. On the other hand, caustic surfaces themselves can by detected using the Lagrangian sub-manifold by identifying the common faces of neighboring tetrahedra 
with opposite volume signs \citep{Shandarin2012}. They are shown in the bottom panel in  \autoref{fig:NstrCaust}. One can see that increasing the refinement factor from 2 to 8 adds mostly walls but the complete wall structure shown in the bottom panel is still considerably greater. Please note that the plots in  two top panels adjusted  exactly to the simulation box in Eulerian space, and the bottom plot shows the Lagrangian box mapped to Eulerian space without adjusting to the simulation box.

There are extensive number of topological indicators in the context of density fields or spatial co-ordinates - such as alpha shapes, Betti numbers, genus statistics. Although a comparative study of these topological measures in multistream fields may be interesting, it is not the intent of this chapter (See Chapter \ref{chapter3b} for a detailed comparison of various structure finding algorithms). In this study, we only investigate percolation transitions in excursion sets of multistreams as a preliminary analysis of topological connectivities. Excursion sets in density fields are shown to have quick percolation transitions \citep{Shandarin2010b} and a similar trend in multistream field is investigated here.

@@

Since the $n_{str}$ field in this study is calculated on regular grids, the boundaries of the structures are not exactly traced. Outlining foldings in the Lagrangian sub-manifolds exactly as shown in \autoref{fig:NstrCaust} or in the {\it flip-flop} calculations shown in \cite{Shandarin2016} give point datasets which are considerably more difficult to analyze. However, recent advancements in computational topology - such as the adaptation of the watershed transforms (using SpineWeb -\citealt{Aragon-Calvo2008} and Morse theory (using DisPerSe - \citealt{Sousbie2011e} and Felix - \citealt{Shivshankar2015a}) to inherently discrete datasets may be useful in the topological analyses of flip-flop fields and caustics.

The multistream field is a scalar function of Eulerian coordinates. We have analyzed functional variation of the $-n_{str}(\bf{x})$ field using Hessian eigenvalues. The Hessian analysis is generally done for inherently continuous fields, For example, Hessian analysis has been previously studied for smoothed density fields (see \citealt{Sousbie2008c}, \citealt{Aragon-Calvo2007}, \citealt{Aragon-Calvo2010}, \citealt{Cautun2014a} etc.), gravitational potential and velocity shear tensor (\citealt{Hoffman2012a}, \citealt{Libeskind2013}, \citealt{Hahn2007}, \citealt{Forero-Romero2009a}, \citealt{Hoffman2012a} and \citealt{Cautun2014a}). Although the multistream field has discrete values by definition, it may be considered smooth for numerical analysis at the scale of grid length of the field. The resulting Hessian eigenvalues characterize the geometry in a four-dimensional hyper-space of ($-n_{str}, x, y, z$). The boundary of a region with $\lambda_1 \geq \lambda_2 \geq \lambda_3 > 0$ is a closed convex contour in this hyper-space, and thus it's projection onto the three-dimensional Lagrangian space is also closed and convex.

@@

Dark matter haloes, being localized structures, are uniquely convenient for our local Hessian analysis. Conditions of $\lambda_1 > 0 > \lambda_2 \geq \lambda_3 $ and $\lambda_1 \geq \lambda_2 > 0 > \lambda_3 $ also give information about curvature. Hessian eigenvalue analysis at high resolution of multistream fields may be very interesting in understanding the tubular edges of filaments and surfaces of walls at smaller scales.

@@

\item We use the multistream field as a proxy for distinguishing of the DM web from DM voids: the web is defined as the regions
with number of streams greater than one and thus voids  as  a single stream regions. The boundary between them  
representing a sharp transition from one-- to three-- stream flow regions
would be a caustic surface in the density field if the mass and spatial resolutions were sufficiently high. They were clearly seen in 
2 simulations by \cite{Melott1989} as well as in three dimensional simulations by \cite{Angulo2016}, \cite{Hahn2016a}, \cite{Hahn2013} and in velocity fields \cite{Hahn2015a}.

@@

The Lagrangian sub-manifold contains dynamical information of structure formation. We analyzed the multistream field that contains the information of foldings in the sub-manifold.  Connectivities in the void and non-void components of the multistream web reveal several details about structure of the Universe that are not probed by traditional density fields. In addition, we demonstrated the use of geometrical features of the multistream field in identifying potential dark matter halo candidates in cosmological N-body simulations.
